PyTracer: Automatically Profiling Numerical Instabilities in Python

### 2022-04-14 IEEEtran/paper-annotated.tex
Revision

\documentclass[11pt]{article}
\usepackage[margin=0.5in]{geometry}
\usepackage{todonotes}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{color}
\usepackage{hyperref}
\usepackage{multirow}
\usepackage{pifont}
\usepackage{xcolor,colortbl}
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=blue!50!black,citecolor=blue!75!black,filecolor=black,urlcolor=blue!75!black}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{xspace}
\usepackage{makecell}
\usepackage{ulem} % provides sout
\usepackage[frozencache=true,cachedir=minted-cache]{minted} 
\usepackage[htt]{hyphenat}      
\usepackage{subcaption}
\usepackage{graphicx,import}
\usepackage{booktabs}

% \usepackage{subfigure}

\newcommand{\tristan}[1]{\color{orange}\textbf{From Tristan:} #1\color{black}\xspace}
\newcommand{\tristanmod}[2]{\color{orange}\sout{#1} #2\color{black}\xspace}
\newcommand{\gkmod}[2]{\color{purple}\sout{#1} #2\color{black}\xspace}

\newcommand{\Yohan}[1]{\color{green!75!black}\textbf{Yohan:} #1\color{black}\xspace}
\newcommand{\cross}[0]{\cellcolor{red!65}\ding{53}}
\newcommand{\valid}[0]{\cellcolor{green!75!black}\ding{51}}
\newcommand{\warn}[0]{\cellcolor{orange!75}?}
\newcommand{\na}[0]{\cellcolor{gray!25}}
\newcommand{\s}[1]{\cellcolor{cyan!25}#1}
\newcommand{\scross}[0]{\ding{53}~}
\newcommand{\svalid}[0]{\ding{51}~}
\newcommand{\swarn}[0]{?~}
\newcommand{\pytracer}[0]{PyTracer\xspace}

@@

\usepackage{tabularx} % provides tabularx to adjust table column sizes to page size
% define thickline for use in tables
\makeatletter
\newcommand{\thickhline}{%
    \noalign {\ifnum 0=`}\fi \hrule height 1pt
    \futurelet \reserved@a \@xhline
}
\newcolumntype{"}{@{\hskip\tabcolsep\vrule width 1pt\hskip\tabcolsep}}
\makeatother

@@

\title{PyTracer: Automatically profiling numerical instabilities in Python}
\author{Yohan Chatelain$^1$, Nigel Young$^1$,  Gregory Kiar$^2$,  Tristan Glatard$^1$\\
$^1$Department of Computer Science and Software Engineering, Concordia University, Montreal, Canada\\
$^2$Center for the Developing Brain, Child Mind Institute, New York, NY, USA}
    \date{}
\maketitle

\begin{abstract}
Numerical stability is a crucial requirement of reliable scientific computing. However, despite the pervasiveness of Python in data science, analyzing large Python programs remains challenging due to the lack of scalable numerical analysis tools available for this language. To fill this gap, we developed \pytracer, a profiler to quantify numerical instability in Python applications. \pytracer transparently instruments Python code to produce numerical traces and visualize them interactively in a Plotly dashboard. We designed \pytracer to be agnostic to numerical noise model , allowing for tool evaluation through Monte-Carlo Arithmetic, random rounding, random data perturbation, or structured noise for a particular application. We illustrate \pytracer's capabilities by testing the numerical stability of key functions in both SciPy and Scikit-learn, two dominant Python libraries for mathematical modeling. Through these evaluations, we demonstrate \pytracer as a scalable, automatic, and generic framework for numerical profiling in Python.
% Numerical stability is a crucial requirement of reliable scientific computing. However, despite the pervasiveness of Python in data science, analyzing large Python programs remains challenging due to the lack of scalable numerical analysis tools available for this language. To address this issue, we developed \pytracer, a profiler to quantify numerical instability in Python applications. \pytracer automatically instruments Python code to produce numerical traces and visualize them interactively in a Plotly dashboard.
% \pytracer is designed to work with any type of numerical noise model, including Monte-Carlo Arithmetic, random rounding or random data perturbations. We illustrate \pytracer's capabilities by testing the numerical stability of key functions in the SciPy and Scikit-learn Python libraries. Finally, \pytracer slows down executions by only $3.5 \times$ on average.
% \tristan{Add one sentence to summarize the findings} 
\end{abstract}

\section{Introduction}

The scientific Python ecosystem has become a central component of data
analyses in recent years, owing to its rich offering of data manipulation, array programming,
numerical analysis, and visualization tools. In particular, libraries such
as NumPy~\cite{harris2020array}, SciPy~\cite{virtanen2020scipy}, scikit-learn~\cite{pedregosa2011scikit} or PyTorch~\cite{paszke2019pytorch} are used in hundreds of publications each year, and serve as a reference set of high-quality open-source core scientific libraries. Researchers have built numerous domain-specific software tools from this ecosystem, leveraging Python's simplicity and flexibility.

@@

We present PyTracer, a numerical stability profiling and visualization tool for Python programs.
\pytracer adopts a dynamic approach that evaluates numerical stability using program execution traces, and is applicable to large programs without manual intervention. The transparent nature of \pytracer is particularly valuable in contrast to theoretical approaches based on condition numbers or backward error analysis, which require detailed modeling of the program and its numerical implementation. Similarly, static code analysis techniques hardly scale to large code bases, particularly in dynamically-typed languages such as Python.

\pytracer estimates the variability of floating-point variables by combining program traces obtained with different numerical perturbations, which requires (1) generating numerically-perturbed executions, (2) tracing floating-point computations along program executions, and (3) visualizing stability evaluations. \pytracer addresses these requirements with (1) a ``fuzzy" Python interpreter instrumented with Monte-Carlo arithmetic, (2) a dynamic instrumentation of Python functions, modules, and classes through meta-programming, (3) an interactive Plotly dashboard to visualize stability measures throughout program executions.

This manuscript presents the design of \pytracer and demonstrates its potential in various use cases. 
Through stability evaluations of the
SciPy and scikit-learn libraries, we demonstrate \pytracer as a practical solution to review the numerical stability of extensive code bases. 
%All information was obtained through \pytracer, visualized separately for improved customizability.
%\tristan{this is a weird place to mention this ^^}

@@

% \subsection{Pytracer workflow}
\pytracer adopts the three-step workflow shown in Figure~\ref{fig:workflow}. First, \pytracer uses \textit{monkey patching} to replace the functions called by the application with a wrapper that saves their arguments and returned values in a trace file.
% \gkmod{}{Q: This is conceptually similar to a decorator, no? If so, perhaps it is worth mentioning} \Yohan{Conceptually, monkey patching is the ability to dynamically change a class or a module. So it's more powerful than the decorators. But yes we use monkey-patching to automatically decorate the functions.}
Second, it executes the application multiple times with noise injection, and aggregates the resulting trace files to compute summary statistics about each traced variable. Finally, a Plotly dashboard provides interactive visualizations highlighting unstable code sections.

@@

\pytracer dynamically instruments Python modules
without requiring source code modifications in the application.
To do so, it creates a new, instrumented version of each module that preserves its original attributes so that the application can transparently use it.
% \gkmod{}{Note: this form of monkey patching is basically global and transparent decorating, right?} \Yohan{Yes}
By default, \pytracer instruments all the attributes
in a given module except the special ones of the form \texttt{\_\_*\_\_}, which is common for reserved Python methods.
However, some attributes are not writable. In this case, \pytracer preservers the original attributes and warns the user. The user can also restrict the list of traced attributes through
an inclusion/exclusion mechanism. The main advantage of this instrumentation approach is the lack of required modification in the application source code, in contrast with decorator-based approaches. In addition, this technique is scalable and applicable in read-only environments such as Singularity containers or HPC clusters.

@@

% As the time of writing this article, fuzzy provides 5 levels of instrumentation (each level N including the lower levels):
% \begin{enumerate}
%     \item BLAS + LAPACK
%     \item Python interpreter
%     \item Numpy
%     \item SciPy
%     \item Scikit-learn
% \end{enumerate}

>>>>

%% bare_jrnl_compsoc.tex
%% V1.4b
%% 2015/08/26
%% by Michael Shell
%% See:
%% http://www.michaelshell.org/
%% for current contact information.
%%
%% This is a skeleton file demonstrating the use of IEEEtran.cls
%% (requires IEEEtran.cls version 1.8b or later) with an IEEE
%% Computer Society journal paper.
%%
%% Support sites:
%% http://www.michaelshell.org/tex/ieeetran/
%% http://www.ctan.org/pkg/ieeetran
%% and
%% http://www.ieee.org/

%%*************************************************************************
%% Legal Notice:
%% This code is offered as-is without any warranty either expressed or
%% implied; without even the implied warranty of MERCHANTABILITY or
%% FITNESS FOR A PARTICULAR PURPOSE! 
%% User assumes all risk.
%% In no event shall the IEEE or any contributor to this code be liable for
%% any damages or losses, including, but not limited to, incidental,
%% consequential, or any other damages, resulting from the use or misuse
%% of any information contained here.
%%
%% All comments are the opinions of their respective authors and are not
%% necessarily endorsed by the IEEE.
%%
%% This work is distributed under the LaTeX Project Public License (LPPL)
%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
%% distributed and modified. A copy of the LPPL, version 1.3, is included
%% in the base LaTeX documentation of all distributions of LaTeX released
%% 2003/12/01 or later.
%% Retain all contribution notices and credits.
%% ** Modified files should be clearly indicated as such, including  **
%% ** renaming them and changing author support contact information. **
%%*************************************************************************

% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
% *** with production work. The IEEE's font choices and paper sizes can   ***
% *** trigger bugs that do not appear when using other class files.       ***                          ***
% The testflow support page is at:
% http://www.michaelshell.org/tex/testflow/

\documentclass[10pt,journal,compsoc]{IEEEtran}

\usepackage{todonotes}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{color}
\usepackage{hyperref}
\usepackage{multirow}
\usepackage{pifont}
\usepackage{xcolor,colortbl}
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=blue!50!black,citecolor=blue!75!black,filecolor=black,urlcolor=blue!75!black}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{xspace}
\usepackage{makecell}
\usepackage{ulem} % provides sout
\usepackage[frozencache=true,cachedir=minted-cache]{minted} 
\usepackage[htt]{hyphenat}      
\usepackage{subcaption}
\usepackage{graphicx,import}
\usepackage{booktabs}
\usepackage{soul}

%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[10pt,journal,compsoc]{../sty/IEEEtran}

% Some very useful LaTeX packages include:
% (uncomment the ones you want to load)

% *** MISC UTILITY PACKAGES ***
%
%\usepackage{ifpdf}
% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
% compilation based on whether the output is pdf or dvi.
% usage:
% \ifpdf
%   % pdf code
% \else
%   % dvi code
% \fi
% The latest version of ifpdf.sty can be obtained from:
% http://www.ctan.org/pkg/ifpdf
% Also, note that IEEEtran.cls V1.7 and later provides a builtin
% \ifCLASSINFOpdf conditional that works the same way.
% When switching from latex to pdflatex and vice-versa, the compiler may
% have to be run twice to clear warning/error messages.

% *** CITATION PACKAGES ***
%
\ifCLASSOPTIONcompsoc
  % IEEE Computer Society needs nocompress option
  % requires cite.sty v4.0 or later (November 2003)
  \usepackage[nocompress]{cite}
\else
  % normal IEEE
  \usepackage{cite}
\fi
% cite.sty was written by Donald Arseneau
% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
% \cite{} output to follow that of the IEEE. Loading the cite package will
% result in citation numbers being automatically sorted and properly
% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
% \cite will automatically add leading space, if needed. Use cite.sty's
% noadjust option (cite.sty V3.8 and later) if you want to turn this off
% such as if a citation ever needs to be enclosed in parenthesis.
% cite.sty is already installed on most LaTeX systems. Be sure and use
% version 5.0 (2009-03-20) and later if using hyperref.sty.
% The latest version can be obtained at:
% http://www.ctan.org/pkg/cite
% The documentation is contained in the cite.sty file itself.
%
% Note that some packages require special options to format as the Computer
% Society requires. In particular, Computer Society  papers do not use
% compressed citation ranges as is done in typical IEEE papers
% (e.g., [1]-[4]). Instead, they list every citation separately in order
% (e.g., [1], [2], [3], [4]). To get the latter we need to load the cite
% package with the nocompress option which is supported by cite.sty v4.0
% and later. Note also the use of a CLASSOPTION conditional provided by
% IEEEtran.cls V1.7 and later.

% *** GRAPHICS RELATED PACKAGES ***
%
\ifCLASSINFOpdf
  % \usepackage[pdftex]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../pdf/}{../jpeg/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\else
  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
  % will default to the driver specified in the system graphics.cfg if no
  % driver is specified.
  % \usepackage[dvips]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../eps/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.eps}
\fi
% graphicx was written by David Carlisle and Sebastian Rahtz. It is
% required if you want graphics, photos, etc. graphicx.sty is already
% installed on most LaTeX systems. The latest version and documentation
% can be obtained at: 
% http://www.ctan.org/pkg/graphicx
% Another good source of documentation is "Using Imported Graphics in
% LaTeX2e" by Keith Reckdahl which can be found at:
% http://www.ctan.org/pkg/epslatex
%
% latex, and pdflatex in dvi mode, support graphics in encapsulated
% postscript (.eps) format. pdflatex in pdf mode supports graphics
% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
% not a bitmapped formats (.jpeg, .png). The IEEE frowns on bitmapped formats
% which can result in "jaggedy"/blurry rendering of lines and letters as
% well as large increases in file sizes.
%
% You can find documentation about the pdfTeX application at:
% http://www.tug.org/applications/pdftex

% *** MATH PACKAGES ***
%
%\usepackage{amsmath}
% A popular package from the American Mathematical Society that provides
% many useful and powerful commands for dealing with mathematics.
%
% Note that the amsmath package sets \interdisplaylinepenalty to 10000
% thus preventing page breaks from occurring within multiline equations. Use:
%\interdisplaylinepenalty=2500
% after loading amsmath to restore such page breaks as IEEEtran.cls normally
% does. amsmath.sty is already installed on most LaTeX systems. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/pkg/amsmath

% *** SPECIALIZED LIST PACKAGES ***
%
%\usepackage{algorithmic}
% algorithmic.sty was written by Peter Williams and Rogerio Brito.
% This package provides an algorithmic environment fo describing algorithms.
% You can use the algorithmic environment in-text or within a figure
% environment to provide for a floating algorithm. Do NOT use the algorithm
% floating environment provided by algorithm.sty (by the same authors) or
% algorithm2e.sty (by Christophe Fiorio) as the IEEE does not use dedicated
% algorithm float types and packages that provide these will not provide
% correct IEEE style captions. The latest version and documentation of
% algorithmic.sty can be obtained at:
% http://www.ctan.org/pkg/algorithms
% Also of interest may be the (relatively newer and more customizable)
% algorithmicx.sty package by Szasz Janos:
% http://www.ctan.org/pkg/algorithmicx

% *** ALIGNMENT PACKAGES ***
%
%\usepackage{array}
% Frank Mittelbach's and David Carlisle's array.sty patches and improves
% the standard LaTeX2e array and tabular environments to provide better
% appearance and additional user controls. As the default LaTeX2e table
% generation code is lacking to the point of almost being broken with
% respect to the quality of the end results, all users are strongly
% advised to use an enhanced (at the very least that provided by array.sty)
% set of table tools. array.sty is already installed on most systems. The
% latest version and documentation can be obtained at:
% http://www.ctan.org/pkg/array

% IEEEtran contains the IEEEeqnarray family of commands that can be used to
% generate multiline equations as well as matrices, tables, etc., of high
% quality.

% *** SUBFIGURE PACKAGES ***
%\ifCLASSOPTIONcompsoc
%  \usepackage[caption=false,font=footnotesize,labelfont=sf,textfont=sf]{subfig}
%\else
%  \usepackage[caption=false,font=footnotesize]{subfig}
%\fi
% subfig.sty, written by Steven Douglas Cochran, is the modern replacement
% for subfigure.sty, the latter of which is no longer maintained and is
% incompatible with some LaTeX packages including fixltx2e. However,
% subfig.sty requires and automatically loads Axel Sommerfeldt's caption.sty
% which will override IEEEtran.cls' handling of captions and this will result
% in non-IEEE style figure/table captions. To prevent this problem, be sure
% and invoke subfig.sty's "caption=false" package option (available since
% subfig.sty version 1.3, 2005/06/28) as this is will preserve IEEEtran.cls
% handling of captions.
% Note that the Computer Society format requires a sans serif font rather
% than the serif font used in traditional IEEE formatting and thus the need
% to invoke different subfig.sty package options depending on whether
% compsoc mode has been enabled.
%
% The latest version and documentation of subfig.sty can be obtained at:
% http://www.ctan.org/pkg/subfig

% *** FLOAT PACKAGES ***
%
%\usepackage{fixltx2e}
% fixltx2e, the successor to the earlier fix2col.sty, was written by
% Frank Mittelbach and David Carlisle. This package corrects a few problems
% in the LaTeX2e kernel, the most notable of which is that in current
% LaTeX2e releases, the ordering of single and double column floats is not
% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
% single column figure to be placed prior to an earlier double column
% figure.
% Be aware that LaTeX2e kernels dated 2015 and later have fixltx2e.sty's
% corrections already built into the system in which case a warning will
% be issued if an attempt is made to load fixltx2e.sty as it is no longer
% needed.
% The latest version and documentation can be found at:
% http://www.ctan.org/pkg/fixltx2e

%\usepackage{stfloats}
% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
% the ability to do double column floats at the bottom of the page as well
% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
% LaTeX2e). It also provides a command:
%\fnbelowfloat
% to enable the placement of footnotes below bottom floats (the standard
% LaTeX2e kernel puts them above bottom floats). This is an invasive package
% which rewrites many portions of the LaTeX2e float routines. It may not work
% with other packages that modify the LaTeX2e float routines. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/pkg/stfloats
% Do not use the stfloats baselinefloat ability as the IEEE does not allow
% \baselineskip to stretch. Authors submitting work to the IEEE should note
% that the IEEE rarely uses double column equations and that authors should try
% to avoid such use. Do not be tempted to use the cuted.sty or midfloat.sty
% packages (also by Sigitas Tolusis) as the IEEE does not format its papers in
% such ways.
% Do not attempt to use stfloats with fixltx2e as they are incompatible.
% Instead, use Morten Hogholm'a dblfloatfix which combines the features
% of both fixltx2e and stfloats:
%
% \usepackage{dblfloatfix}
% The latest version can be found at:
% http://www.ctan.org/pkg/dblfloatfix

%\ifCLASSOPTIONcaptionsoff
%  \usepackage[nomarkers]{endfloat}
% \let\MYoriglatexcaption\caption
% \renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
%\fi
% endfloat.sty was written by James Darrell McCauley, Jeff Goldberg and 
% Axel Sommerfeldt. This package may be useful when used in conjunction with 
% IEEEtran.cls'  captionsoff option. Some IEEE journals/societies require that
% submissions have lists of figures/tables at the end of the paper and that
% figures/tables without any captions are placed on a page by themselves at
% the end of the document. If needed, the draftcls IEEEtran class option or
% \CLASSINPUTbaselinestretch interface can be used to increase the line
% spacing as well. Be sure and use the nomarkers option of endfloat to
% prevent endfloat from "marking" where the figures would have been placed
% in the text. The two hack lines of code above are a slight modification of
% that suggested by in the endfloat docs (section 8.4.1) to ensure that
% the full captions always appear in the list of figures/tables - even if
% the user used the short optional argument of \caption[]{}.
% IEEE papers do not typically make use of \caption[]'s optional argument,
% so this should not be an issue. A similar trick can be used to disable
% captions of packages such as subfig.sty that lack options to turn off
% the subcaptions:
% For subfig.sty:
% \let\MYorigsubfloat\subfloat
% \renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}}
% However, the above trick will not work if both optional arguments of
% the \subfloat command are used. Furthermore, there needs to be a
% description of each subfigure *somewhere* and endfloat does not add
% subfigure captions to its list of figures. Thus, the best approach is to
% avoid the use of subfigure captions (many IEEE journals avoid them anyway)
% and instead reference/explain all the subfigures within the main caption.
% The latest version of endfloat.sty and its documentation can obtained at:
% http://www.ctan.org/pkg/endfloat
%
% The IEEEtran \ifCLASSOPTIONcaptionsoff conditional can also be used
% later in the document, say, to conditionally put the References on a 
% page by themselves.

% *** PDF, URL AND HYPERLINK PACKAGES ***
%
%\usepackage{url}
% url.sty was written by Donald Arseneau. It provides better support for
% handling and breaking URLs. url.sty is already installed on most LaTeX
% systems. The latest version and documentation can be obtained at:
% http://www.ctan.org/pkg/url
% Basically, \url{my_url_here}.

% *** Do not adjust lengths that control margins, column widths, etc. ***
% *** Do not use packages that alter fonts (such as pslatex).         ***
% There should be no need to do such things with IEEEtran.cls V1.6 and later.
% (Unless specifically asked to do so by the journal or conference you plan
% to submit to, of course. )

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}

\newcommand{\cross}[0]{\cellcolor{red!65}\ding{53}}
\newcommand{\valid}[0]{\cellcolor{green!75!black}\ding{51}}
\newcommand{\warn}[0]{\cellcolor{orange!75}?}
\newcommand{\na}[0]{\cellcolor{gray!25}}
\newcommand{\s}[1]{\cellcolor{cyan!25}#1}
\newcommand{\scross}[0]{\ding{53}~}
\newcommand{\svalid}[0]{\ding{51}~}
\newcommand{\swarn}[0]{?~}
\newcommand{\pytracer}[0]{PyTracer\xspace}

\definecolor{lightred}{rgb}{1.0,0.6,0.6}
\definecolor{lightgreen}{rgb}{0.6,1.0,0.6}

\DeclareRobustCommand{\add}[1]{{\sethlcolor{lightgreen}\hl{#1}}}
\DeclareRobustCommand{\remove}[1]{{\sethlcolor{lightred}\hl{#1}}}

\DeclareRobustCommand{\addCite}[1]{{\sethlcolor{lightgreen}\hl{\mbox{#1}}}}
\DeclareRobustCommand{\removeCite}[1]{{\sethlcolor{lightred}\hl{\mbox{#1}}}}

@@

\makeatletter
\newcommand{\thickhline}{%
    \noalign {\ifnum 0=`}\fi \hrule height 1pt
    \futurelet \reserved@a \@xhline
}
\newcolumntype{"}{@{\hskip\tabcolsep\vrule width 1pt\hskip\tabcolsep}}
\makeatother

@@

%
% paper title
% Titles are generally capitalized except for words such as a, an, and, as,
% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
% not capitalized unless they are the first or last word of the title.
% Linebreaks \\ can be used within to get better formatting as desired.
% Do not put math or special symbols in the title.
\title{PyTracer: Automatically Profiling Numerical Instabilities in Python}
%
%
% author names and IEEE memberships
% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
% a structure at a ~ so this keeps an author's name from being broken across
% two lines.
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
%
%
%\IEEEcompsocitemizethanks is a special \thanks that produces the bulleted
% lists the Computer Society journals use for "first footnote" author
% affiliations. Use \IEEEcompsocthanksitem which works much like \item
% for each affiliation group. When not in compsoc mode,
% \IEEEcompsocitemizethanks becomes like \thanks and
% \IEEEcompsocthanksitem becomes a line break with idention. This
% facilitates dual compilation, although admittedly the differences in the
% desired content of \author between the different types of papers makes a
% one-size-fits-all approach a daunting prospect. For instance, compsoc 
% journal papers have the author affiliations above the "Manuscript
% received ..."  text while in non-compsoc journals this is reversed. Sigh.

\author{Yohan Chatelain$^1$, Nigel Yong Sao Young$^1$,  Gregory Kiar$^2$,  Tristan Glatard$^1$\\
    $^1$Department of Computer Science and Software Engineering, Concordia University, Montreal, Canada\\
    $^2$Center for the Developing Brain, Child Mind Institute, New York, NY, USA}
%% \author{Michael~Shell,~\IEEEmembership{Member,~IEEE,}
%%         John~Doe,~\IEEEmembership{Fellow,~OSA,}
%%         and~Jane~Doe,~\IEEEmembership{Life~Fellow,~IEEE}% <-this % stops a space
%% \IEEEcompsocitemizethanks{\IEEEcompsocthanksitem M. Shell was with the Department
%% of Electrical and Computer Engineering, Georgia Institute of Technology, Atlanta,
%% GA, 30332.\protect\\
%% % note need leading \protect in front of \\ to get a newline within \thanks as
%% % \\ is fragile and will error, could use \hfil\break instead.
%% E-mail: see http://www.michaelshell.org/contact.html
%% \IEEEcompsocthanksitem J. Doe and J. Doe are with Anonymous University.}% <-this % stops an unwanted space
%% \thanks{Manuscript received April 19, 2005; revised August 26, 2015.}}

% note the % following the last \IEEEmembership and also \thanks - 
% these prevent an unwanted space from occurring between the last author name
% and the end of the author line. i.e., if you had this:
% 
% \author{....lastname \thanks{...} \thanks{...} }
%                     ^------------^------------^----Do not want these spaces!
%
% a space would be appended to the last name and could cause every name on that
% line to be shifted left slightly. This is one of those "LaTeX things". For
% instance, "\textbf{A} \textbf{B}" will typeset as "A B" not "AB". To get
% "AB" then you have to do: "\textbf{A}\textbf{B}"
% \thanks is no different in this regard, so shield the last } of each \thanks
% that ends a line with a % and do not let a space in before the next \thanks.
% Spaces after \IEEEmembership other than the last one are OK (and needed) as
% you are supposed to have spaces between the names. For what it is worth,
% this is a minor point as most people would not even notice if the said evil
% space somehow managed to creep in.

% The paper headers
\markboth{Journal of \LaTeX\ Class Files,~Vol.~14, No.~8, August~2015}%
{Shell \MakeLowercase{\textit{et al.}}: Bare Demo of IEEEtran.cls for Computer Society Journals}
% The only time the second header will appear is for the odd numbered pages
% after the title page when using the twoside option.
% 
% *** Note that you probably will NOT want to include the author's ***
% *** name in the headers of peer review papers.                   ***
% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
% you desire.

% The publisher's ID mark at the bottom of the page is less important with
% Computer Society journal papers as those publications place the marks
% outside of the main text columns and, therefore, unlike regular IEEE
% journals, the available text space is not reduced by their presence.
% If you want to put a publisher's ID mark on the page you can do it like
% this:
%\IEEEpubid{0000--0000/00\$00.00~\copyright~2015 IEEE}
% or like this to get the Computer Society new two part style.
%\IEEEpubid{\makebox[\columnwidth]{\hfill 0000--0000/00/\$00.00~\copyright~2015 IEEE}%
%\hspace{\columnsep}\makebox[\columnwidth]{Published by the IEEE Computer Society\hfill}}
% Remember, if you use this you must call \IEEEpubidadjcol in the second
% column for its text to clear the IEEEpubid mark (Computer Society jorunal
% papers don't need this extra clearance.)

% use for special paper notices
%\IEEEspecialpapernotice{(Invited Paper)}

% for Computer Society papers, we must declare the abstract and index terms
% PRIOR to the title within the \IEEEtitleabstractindextext IEEEtran
% command as these need to go into the title area created by \maketitle.
% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\IEEEtitleabstractindextext{%
    \begin{abstract}
        Numerical stability is a crucial requirement of reliable scientific
        computing. However, despite the pervasiveness of Python in data science,
        analyzing large Python programs remains challenging due to the lack of
        scalable numerical analysis tools available for this language. To fill
        this gap, we developed \pytracer, a profiler to quantify numerical
        instability in Python applications. \pytracer transparently instruments
        Python code to produce numerical traces and visualize them interactively
        in a Plotly dashboard. We designed \pytracer to be agnostic to numerical
        noise model , allowing for tool evaluation through Monte-Carlo
        Arithmetic, random rounding, random data perturbation, or structured
        noise for a particular application. We illustrate \pytracer's
        capabilities by testing the numerical stability of key functions in both
        SciPy and Scikit-learn, two dominant Python libraries for mathematical
        modeling. Through these evaluations, we demonstrate \pytracer as a
        scalable, automatic, and generic framework for numerical profiling in
        Python.
    \end{abstract}
}
% Note that keywords are not normally used for peerreview papers.
%% \begin{IEEEkeywords}
%% Computer Society, IEEE, IEEEtran, journal, \LaTeX, paper, template.
%% \end{IEEEkeywords}}

% make the title area
\maketitle

% To allow for easy dual compilation without having to reenter the
% abstract/keywords data, the \IEEEtitleabstractindextext text will
% not be used in maketitle, but will appear (i.e., to be "transported")
% here as \IEEEdisplaynontitleabstractindextext when the compsoc 
% or transmag modes are not selected <OR> if conference mode is selected 
% - because all conference papers position the abstract like regular
% papers do.
\IEEEdisplaynontitleabstractindextext
% \IEEEdisplaynontitleabstractindextext has no effect when using
% compsoc or transmag under a non-conference mode.

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle

\IEEEraisesectionheading{\section{Introduction}\label{sec:introduction}}
% Computer Society journal (but not conference!) papers do something unusual
% with the very first section heading (almost always called "Introduction").
% They place it ABOVE the main text! IEEEtran.cls does not automatically do
% this for you, but you can achieve this effect with the provided
% \IEEEraisesectionheading{} command. Note the need to keep any \label that
% is to refer to the section immediately after \section in the above as
% \IEEEraisesectionheading puts \section within a raised box.

\IEEEPARstart{T}{he} scientific Python ecosystem has become a central component
of data analyses in recent years, owing to its rich offering of data
manipulation, array programming, numerical analysis, and visualization tools. In
particular, libraries such as NumPy~\cite{harris2020array},
SciPy~\cite{virtanen2020scipy}, scikit-learn~\cite{pedregosa2011scikit} or
PyTorch~\cite{paszke2019pytorch} are used in hundreds of publications each year,
and serve as a reference set of high-quality open-source core scientific
libraries. Researchers have built numerous domain-specific software tools from
this ecosystem, leveraging Python's simplicity and flexibility.

@@

We present PyTracer, a numerical stability profiling and visualization tool for
Python programs. \pytracer adopts a dynamic approach that evaluates numerical
stability using program execution traces, and is applicable to large programs
without manual intervention. The transparent nature of \pytracer is particularly
valuable in contrast to theoretical approaches based on condition numbers or
backward error analysis~\addCite{\cite{higham2002accuracy}}, which require detailed modeling of the program and its
numerical implementation. Similarly, static code analysis techniques~\addCite{\cite{goubault2006static}} hardly
scale to large code bases, particularly in dynamically-typed languages such as
Python.

\pytracer estimates the variability of floating-point variables by combining
program traces obtained with different numerical perturbations, which requires
(\add{a}\remove{1}) generating numerically-perturbed executions, (\add{b}\remove{2}) tracing floating-point
computations along program executions, and (\add{c}\remove{3}) visualizing stability
evaluations. 
\add{PyTracer uses the ``fuzzy" Python environment instrumented with Monte-Carlo
arithmetic to address the requirement (a). \mbox{\pytracer} addresses the
remaining requirements with (b) a dynamic instrumentation of Python functions,
modules, and classes through meta-programming, (c) an interactive Plotly
dashboard to visualize stability measures throughout program executions.
\mbox{\pytracer} is available at
\mbox{\href{https://github.com/big-data-lab-team/pytracer.git}{github.com/big-data-lab-team/pytracer.git}}
under MIT License.}
\remove{\mbox{\pytracer} addresses these requirements with (1) a ``fuzzy" Python
    interpreter instrumented with Monte-Carlo arithmetic, (2) a dynamic
    instrumentation of Python functions, modules, and classes through
    meta-programming, (3) an interactive Plotly dashboard to visualize stability
    measures throughout program executions. }

This manuscript presents the design of \pytracer and demonstrates its potential
in various use cases. Through stability evaluations of the SciPy and
scikit-learn libraries, we demonstrate \pytracer as a practical solution to
review the numerical stability of extensive code bases.

@@

% \subsection{Pytracer workflow}
\pytracer adopts the three-step workflow shown in Figure~\ref{fig:workflow}.
First, \pytracer uses \textit{monkey patching} to replace the functions called
by the application with a wrapper that saves their arguments and returned values
in a trace file.
Second, it executes the application multiple times with noise injection, and
aggregates the resulting trace files to compute summary statistics about each
traced variable. Finally, a Plotly dashboard provides interactive visualizations
highlighting unstable code sections.

@@

\pytracer dynamically instruments Python modules without requiring source code
modifications in the application. To do so, it creates a new, instrumented
version of each module that preserves its original attributes so that the
application can transparently use it.
% \gkmod{}{Note: this form of monkey patching is basically global and
% transparent decorating, right?} \Yohan{Yes}
By default, \pytracer instruments all the attributes in a given module except
the special ones of the form \texttt{\_\_*\_\_}, which is common for reserved
Python methods. However, some attributes are not writable. In this case,
\pytracer \remove{preservers} \add{preserves} the original attributes and warns the user. The user can
also restrict the list of traced attributes through an inclusion/exclusion
mechanism. The main advantage of this instrumentation approach is the lack of
required modification in the application source code, in contrast with
\remove{decorator-based approaches} \add{a decorator-based approache~\mbox{\cite{hunt2019decorators}}}. 
In addition, this technique is scalable and
applicable in read-only environments such as Singularity containers or HPC
clusters.

@@
